\begin{acknowledgements}
    I'd like to start by thanking professor Viroli and his research team, in particular Gianluca, Roberto and Danilo, for giving me the opportunity to work on this thesis and for their constant presence in the past months.
    %
    I really hope to become a part of this reality after graduation.

    Next comes my family, for the unconditional support they have been giving me throughout my studies, for always cheering with joy for my successes, and for understanding the effort I've been putting in during this period.
    %
    I consider myself really lucky for this.

    Thanks to my closest friends, Agnese, Bruno, Gianluca, and Ylenia, with whom I bonded so much over the past few years.
    %
    I am immensely grateful that you are part of my life and I hope that you'll all forgive me for being less present in the last period.
    %
    I do plan to make a comeback, though, so you can rest assured.

    A huge thanks to all the past, present, and future members of S.P.R.I.Te.
    %
    This association is what brings joy and excitement to the community of students in Cesena, helping them consider the academic journey a pleasure more than a duty.
    %
    Being part of this changed me a lot and for the better, while also helping me come in touch with different aspects of university that I would have never had a chance to know, so I am truly thankful.
    
    Thanks to all my groups of friends: \quotes{Gli amici di Sprite}, \quotes{Gruppo Badia}, \quotes{Party Hard @ Cesena}, and \quotes{Lo Spaceteam}.
    %
    Hanging out with all of you is always a pleasure that helps me recharge my batteries when I feel low.

    Thanks to my colleagues, but most importantly friends, from EasyDesk for the experiences we shared together.
    %
    I'm sure the next few years will be full of gratification and fulfillment, because we all know the hard work we put in.

    Finally, thanks to my girlfriend Lucrezia, for being the wonderful person she is.
    %
    You've given me strength, happiness, and love, helping me a lot through one of the most stressful moments of my life.
    %
    I hope that you can see how much this means to me and that someday I'll be able to do the same to you.
\end{acknowledgements}